---
title: "OD Bath"
author: "Blaise Kelly"
date: "29/05/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(sf)
library(mapview)
library(lubridate)
library(stringr)
library(reshape2)
library(knitr)
library(tmap)
library(leaflet)
library(osrm)

##define coordinate systems
latlong = "+init=epsg:4326"
ukgrid = "+init=epsg:27700"

load("out/pct_od.RData")
load("out/jittered_routes.RData")
load("out/validation_dat.RData")

```

## Introduction

Data from the 2011 UK census data is queried using the Propensity to Cycle Tool (pct) R package. Origin Destination (OD) matrices are created and 'jittered' using the ODjitter R package.

## area of interest

```{r area_of_interest, echo = FALSE, warning = FALSE}

zones_cent <- st_centroid(zones_in)

## pick out example zone
zone_A <- filter(zones_in, geo_code == "E01014479")
## extract centre point
zone_A_cent <- st_centroid(zone_A)
## extracts roads
zone_rds_A <- st_intersection(rdz, zone_A)
## pick out exanmple second zone
zone_B <- filter(zones_in, geo_code == "E01014482")
## extrract centre point
zone_B_cent <- st_centroid(zone_B)
## roads for zone B
zone_rds_B <- st_intersection(rdz, zone_B) %>% 
  select(osm_id)

LSOA_small <- filter(LSOA_in, geo_code1 %in% c("E01014479", "E01014482") & geo_code2 == c("E01014479", "E01014482"))

zones_small <- filter(zones_in, geo_code == "E01014482" | geo_code == "E01014479")

desire_lines_small = stplanr::od2line(flow = LSOA_small,zones = zones_small)

## show buildings as example of where people live, but also as possiblity for jittering
buildings_in <- buildings[zone_A,]

lon <- st_coordinates(st_centroid(vgt_area))[1]
lat <- st_coordinates(st_centroid(vgt_area))[2]

m <- leaflet() %>% 
  setView(lon, lat, zoom = 11) %>% 
  addProviderTiles('CartoDB.Positron')

m <- m %>% addPolygons(data = vgt_area, color = 'black', weight = 1, fillColor = "grey",
                       opacity = 0.5, fillOpacity = 0.5,
                       highlightOptions = highlightOptions(color = "white", weight = 2,
                                                           bringToFront = FALSE), group = "area of interest")

m <- m %>% addLayersControl(overlayGroups = c("area of interest"),
                            options = layersControlOptions(collapsed = FALSE), position = "topright")


m

```

Import the census zones from the pct data from pct package and intersect with area of interest


```{r zones_in, echo = FALSE, warning = FALSE}

m <- leaflet() %>% 
  setView(lon, lat, zoom = 11) %>% 
  addProviderTiles('CartoDB.Positron')

m <- m %>% addPolygons(data = zones_in, color = 'black', weight = 1, fillColor = "yellow",
                       opacity = 1.0, fillOpacity = 0.5,
                       highlightOptions = highlightOptions(color = "white", weight = 2,
                                                           bringToFront = FALSE), group = "zones")

m <- m %>% addCircleMarkers(data = zones_cent, color = "black", fillColor = "black", weight = 2, radius = 1,
                            group = "zone centre points")


m <- m %>% addLayersControl(overlayGroups = c("zone", "zone centre points"),
                            options = layersControlOptions(collapsed = FALSE), position = "topright")


m


```

Download Origin Destination (OD) data for zones from LSOA data store. Generate OD desire lines. 

```{r dls, echo=FALSE, warning=FALSE}

me_pal <- c("#0000b3", "#0000eb", "#1d00ff", "#4a00ff", "#7600ff", "#a211ee", "#cf2ed1", "#fb4ab5", 
            "#ff6798", "#ff837c", "#ff9f60", "#ffbc43", "#ffd827", "#fff50a")

desire_lines_motor <- select(desire_lines, car_driver, motorbike, bus, taxi_other)

vehicle_types <- c("car_driver", "motorbike", "bus", "taxi_other")

max_n <- max(desire_lines$car_driver, desire_lines$motorbike, desire_lines$bus, desire_lines$taxi_other)

pal_v <- colorFactor(me_pal, domain = seq(0,max_n, length.out = 10), reverse = FALSE)

m <- leaflet() %>% 
  setView(lon, lat, zoom = 11) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addScaleBar(position = "bottomleft", options = scaleBarOptions(metric = TRUE))

m <- m %>% addPolygons(data = zones_in, color = 'black', weight = 1, fillColor = "yellow",
                       opacity = 1.0, fillOpacity = 0.5,
                       highlightOptions = highlightOptions(color = "white", weight = 2,
                                                           bringToFront = FALSE), group = "zones")

for (v in vehicle_types){

  v_df <- desire_lines_motor %>% 
    select(vtype = v) %>% 
    filter(vtype > 0)
  
  v_df$vtype <- as.numeric(v_df$vtype)
  
m <- m %>% addPolylines(data = v_df, color = ~pal_v(v_df$vtype), weight = 2, group = v)

}

m <- m %>% addLegend("bottomleft", pal=pal_v, values=seq(0,max_n, length.out = 10), opacity=1, title = "trips")

m <- m %>% addLayersControl(overlayGroups = c("zones"), baseGroups = vehicle_types,
                            options = layersControlOptions(collapsed = FALSE), position = "topright")


m

```

Using the osrm R package an estimated route is generated for each car unique OD combination and the journeys assigned. The routes are then combined using the overline function from the stplanr R package

```{r overline, echo = FALSE, warning = FALSE}

max_cars <- max(r$car_driver)

bks_hr <- seq(from = 0, to = max(r$car_driver), by = max_cars/NROW(me_pal))

# tmaptools::palette_explorer()
tm0 <- tm_shape(r) +
  tm_lines(palette = me_pal, breaks = bks_hr,
           lwd = "car_driver",
           scale = 20,
           title.lwd = "Number of vehicles",
           alpha = 0.6,
           col = "car_driver",
           title = "Flow (number of vehicles)"
  ) +
  tm_layout(frame = FALSE, legend.outside = TRUE)+
  tm_scale_bar(position = c("left", "bottom"))

tm0

```

Assuming people start from the centre of a zone can seem like an insignificant simplification when looking at a large area, however, looking more closely at the possibilities for start and end points within a zone and the effect they can have shows it can make a big difference to the likely route someone will take.

Using zone E01014479, which will be called 'zone A' as an example below, every journey could theoretically start or finish at any point in the zone. However, even if someone starts from the middle of a field they will not be able to begin driving until they reach a road. Similarly it is also quite likely journeys will start and end at buildings. For privacy and data simplification reasons the census data is not available at household level, so knowing where in a zone a journey starts or ends is not possible.

```{r zones_buildings, echo = FALSE, warning = FALSE}

pal_B <- colorFactor("viridis", domain = buildings_in$building, reverse = FALSE)

lon <- st_coordinates(zone_A_cent)[,1]
lat <- st_coordinates(zone_A_cent)[,2]

m <- leaflet() %>% 
  setView(lon, lat, zoom = 15) %>% 
  addProviderTiles('CartoDB.Positron')

m <- m %>% addPolygons(data = buildings_in, color = 'black', weight = 1, fillColor = ~pal_B(buildings_in$building),
                       opacity = 1.0, fillOpacity = 1,
                       highlightOptions = highlightOptions(color = "white", weight = 2,
                                                           bringToFront = FALSE), group = "buildings")

m <- m %>% addPolygons(data = zone_A, color = 'black', weight = 1, fillColor = "grey",
                       opacity = 1.0, fillOpacity = 0.2,
                       highlightOptions = highlightOptions(color = "white", weight = 2,
                                                           bringToFront = FALSE), group = "zone")

m <- m %>% addCircleMarkers(data = zone_A_cent, color = "red", weight = 2, group = "zone_centre")

m <- m %>% addPolylines(data = zone_rds_A, color = "purple", weight = 2, group = "roads")

m <- m %>% addLegend("bottomleft", pal=pal_B, values=buildings_in$building, opacity=1, title = "Type",
                     group = "Buildings")

m <- m %>% addLayersControl(overlayGroups = c("buildings", "roads", "zone_centre", "zone"),
                            options = layersControlOptions(collapsed = FALSE), position = "topright")


m


```

Taking zone E01014479 as 'zone A' and E01014482 as 'zone B' as example OD pairs. The possible start and end points, if the road network is split into 5 metre uniform points are shown below.




Generating routes between these different pairs the difference start and finish point makes on a journey is shown below, by selecting 4 different route options.

```{r variation, echo = FALSE, warning = FALSE}

baf <- osmdata::getbb("Bath", format_out = "sf_polygon") %>% 
  st_transform(ukgrid) %>% 
  st_buffer(-200) %>% 
  st_transform(latlong)

baf_zones <- zones_in[baf,]


approx_interval <- 5

vz <- unique(zone_rds_B$osm_id)

vgt_split <- list()
for (v in vz){
  tryCatch({
    df <- zone_rds_B %>% filter(osm_id == v) %>% st_transform(ukgrid) %>% mutate(lngths = as.numeric(st_length(geometry)))
    
    n_pts <- df$lngths/approx_interval
    df1 <- st_line_sample(df, sample = 0)
    df2 <- st_line_sample(df, n = n_pts)
    df3 <- st_line_sample(df, sample = 1)
    df1 <- data.frame(st_coordinates(df1))
    df2 <- data.frame(st_coordinates(df2))
    df3 <- data.frame(st_coordinates(df3))
    df <- rbind(df1, df2, df3)
    df <- transmute(df, Source.name = v, X, Y)
    vgt_split[[v]] <- df
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

vgt <- do.call(rbind, vgt_split)

B_pts <- st_as_sf(vgt, coords = c("X", "Y"), crs = ukgrid)
B_pts$ID <- rownames(B_pts)
B_pts$zone <- "zone_B"

vz <- unique(zone_rds_A$osm_id)

vgt_split <- list()
for (v in vz){
  tryCatch({
    df <- zone_rds_A %>% filter(osm_id == v) %>% st_transform(ukgrid) %>% mutate(lngths = as.numeric(st_length(geometry)))
    
    n_pts <- df$lngths/approx_interval
    df1 <- st_line_sample(df, sample = 0)
    df2 <- st_line_sample(df, n = n_pts)
    df3 <- st_line_sample(df, sample = 1)
    df1 <- data.frame(st_coordinates(df1))
    df2 <- data.frame(st_coordinates(df2))
    df3 <- data.frame(st_coordinates(df3))
    df <- rbind(df1, df2, df3)
    df <- transmute(df, Source.name = v, X, Y)
    vgt_split[[v]] <- df
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

vgt <- do.call(rbind, vgt_split)

A_pts <- st_as_sf(vgt, coords = c("X", "Y"), crs = ukgrid)
A_pts$ID <- rownames(A_pts)
A_pts$zone <- "zone_A"

pts <- rbind(A_pts, B_pts)

## examples of opposite jounneys

route_pairs <- data.frame(start = c("121120357.6",
                                    "671912551.2",
                                    "671912551.2",
                                    "510800084.56"),
                          end = c("510800084.56", 
                                  "89126716.40", 
                                  "121120357.6", 
                                  "89126716.40"))

routez <- list()
starts <- list()
ends <- list()
for (r in seq(1:NROW(route_pairs))){
  
  rp <- route_pairs[r,]
  
  sp <- pts %>% filter(ID == rp$start) %>% st_transform(latlong)
  ep <- pts %>% filter(ID == rp$end) %>% st_transform(latlong)
  
  route <- osrmRoute(src = sp, dst = ep,
                     overview = "full", returnclass = "sf")
  
  nam <- as.character(r)
  routez[[nam]] <- route
  starts[[nam]] <- sp
  ends[[nam]] <- ep
  
}

all_routes <- do.call(rbind, routez)
all_starts <- do.call(rbind, starts)
all_ends <- do.call(rbind, ends)

lon <- mean(st_coordinates(baf_zones)[,1])
lat <- mean(st_coordinates(baf_zones)[,2])

m <- leaflet() %>% 
  setView(lon, lat, zoom = 13) %>% 
  addProviderTiles('CartoDB.Positron')

m <- m %>% addPolygons(data = zone_B, color = 'black', weight = 1, fillColor = "grey",
                       opacity = 1.0, fillOpacity = 0.2,
                       highlightOptions = highlightOptions(color = "white", weight = 2,
                                                           bringToFront = FALSE), group = "zones")

m <- m %>% addPolygons(data = zone_A, color = 'black', weight = 1, fillColor = "grey",
                       opacity = 1.0, fillOpacity = 0.2,
                       highlightOptions = highlightOptions(color = "white", weight = 2,
                                                           bringToFront = FALSE), group = "zones")

m <- m %>% addPolylines(data = zone_rds_A, color = "yellow", weight = 2, group = "osm road geometry")

m <- m %>% addPolylines(data = zone_rds_B, color = "yellow", weight = 2, group = "osm road geometry")

m <- m %>% addCircleMarkers(data = A_pts, color = "red", fillColor = "red", weight = 2, radius = 2, group = "points")

m <- m %>% addCircleMarkers(data = B_pts, color = "red", fillColor = "red", weight = 2, radius = 2, group = "points")

m <- m %>% addLayersControl(overlayGroups = c("zones", "osm road geometry", "points"),
                            options = layersControlOptions(collapsed = FALSE), position = "topright")

m

```

```{r bath, echo = FALSE, warning = FALSE}


LSOA_small <- filter(LSOA_in, geo_code1 %in% c("E01014479", "E01014482") & geo_code2 == c("E01014479", "E01014482"))

zones_small <- filter(zones_in, geo_code == "E01014482" | geo_code == "E01014479")

zones_cent <- st_centroid(zones_small)

route_small <- osrmRoute(src = zones_cent[1,], dst = zones_cent[2,],
                   overview = "full", returnclass = "sf")

lon <- st_coordinates(st_centroid(baf[1,]))[1]
lat <- st_coordinates(st_centroid(baf[1,]))[2]

m <- leaflet() %>% 
  setView(lon, lat, zoom = 13) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addScaleBar(position = "bottomleft", options = scaleBarOptions(metric = TRUE))

m <- m %>% addPolygons(data = zones_small, color = 'green', weight = 2, fillColor = "green",
                       opacity = 1.0, fillOpacity = 0.3,
                       highlightOptions = highlightOptions(color = "white", weight = 2,
                                                           bringToFront = FALSE), group = "zones A/B")

m <- m %>% addPolygons(data = baf_zones, color = 'black', weight = 1, fillColor = "grey",
                       opacity = 1.0, fillOpacity = 0.2,
                       highlightOptions = highlightOptions(color = "white", weight = 2,
                                                           bringToFront = FALSE), group = "bath zones")

m <- m %>% addCircleMarkers(data = zones_cent, color = "red", weight = 2, group = "zone_centre")

m <- m %>% addPolylines(data = desire_lines_small, color = "purple", weight = 2, group = "desire line")


m <- m %>% addLayersControl(overlayGroups = c("zones A/B", "bath zones", "zone_centre", "desire line"),
                            options = layersControlOptions(collapsed = FALSE), position = "topright")


m
 

```


```{r start_end, echo = FALSE, warning = FALSE}


m <- leaflet() %>% 
  setView(lon, lat, zoom = 14) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addScaleBar(position = "bottomleft", options = scaleBarOptions(metric = TRUE))

m <- m %>% addPolygons(data = zones_small, color = 'green', weight = 2, fillColor = "green",
                       opacity = 1.0, fillOpacity = 0.3,
                       highlightOptions = highlightOptions(color = "white", weight = 2,
                                                           bringToFront = FALSE), group = "zones small")

m <- m %>% addPolygons(data = baf_zones, color = 'black', weight = 1, fillColor = "grey",
                       opacity = 1.0, fillOpacity = 0.2,
                       highlightOptions = highlightOptions(color = "white", weight = 2,
                                                           bringToFront = FALSE), group = "bath zones")

m <- m %>% addCircleMarkers(data = zones_cent, color = "red", weight = 2, group = "route centre")

m <- m %>% addPolylines(data = route_small, color = "purple", weight = 2, group = "route centre")

m <- m %>% addPolylines(data = all_routes[1,], color = "purple", weight = 2, group = "route 1")

m <- m %>% addCircleMarkers(data = all_starts[1,], color = "green", weight = 2, group = "route 1")

m <- m %>% addCircleMarkers(data = all_ends[1,], color = "red", weight = 2, group = "route 1")

m <- m %>% addPolylines(data = all_routes[2,], color = "purple", weight = 2, group = "route 2")

m <- m %>% addCircleMarkers(data = all_starts[2,], color = "green", weight = 2, group = "route 2")

m <- m %>% addCircleMarkers(data = all_ends[2,], color = "red", weight = 2, group = "route 2")

m <- m %>% addPolylines(data = all_routes[3,], color = "purple", weight = 2, group = "route 3")

m <- m %>% addCircleMarkers(data = all_starts[3,], color = "green", weight = 2, group = "route 3")

m <- m %>% addCircleMarkers(data = all_ends[3,], color = "red", weight = 2, group = "route 3")

m <- m %>% addPolylines(data = all_routes[4,], color = "purple", weight = 2, group = "route 4")

m <- m %>% addCircleMarkers(data = all_starts[4,], color = "green", weight = 2, group = "route 4")

m <- m %>% addCircleMarkers(data = all_ends[4,], color = "red", weight = 2, group = "route 4")

m <- m %>% addLayersControl(overlayGroups = c("zones small", "bath zones"),
                            baseGroups = c("route 1", "route 2", "route 3", "route 4", "route centre"),
                            options = layersControlOptions(collapsed = FALSE), position = "topright")


m

```

Using the methodology OD jitter developed by ... the OD start and end points can be randomly varied based on a defined geometry within each zone. In this case the osm road geometry was used as previously described.

[](out/plots/OD_raw.gif)

[](out/plots/OD_jitter.gif)

[](out/plots/OD_jitter_trimmed.gif)

In 2017 two detailed surveys were undertaken in the Bath area. One using pneumatic loops to count vehicles passing and another using ANPR data.

Both these datasets can be used to validate the estimates generated from the OD matrices.

```{r ol_routes, echo=FALSE, warning=FALSE}

library(stplanr)

site_coords <- st_coordinates(all_sites)

bb <- st_as_sfc(st_bbox(c(xmin = min(site_coords[,1]), xmax = max(site_coords[,1]), ymax = max(site_coords[,2]), ymin = min(site_coords[,2])), crs = st_crs(4326)))

bb_buff <- bb %>% 
  st_transform(ukgrid) %>% 
  st_buffer(1000) %>% 
  st_transform(latlong)

# snap the atcs locations to the roads
rdz_noresi <- rdz[bb_buff,] %>% 
  select(osm_id, geometry, highway) %>% 
  filter(highway %in% c("tertiary", "primary", "trunk")) %>% 
  st_transform(ukgrid) %>% 
  st_line_sample(n = 20) %>% 
  st_cast("POINT") %>% 
st_transform(latlong) %>% 
  st_as_sf()

rdz_resi <- rdz[bb_buff,] %>% 
  select(osm_id, geometry, highway) %>% 
  filter(highway %in% c("residential")) %>% 
  st_transform(ukgrid) %>% 
  st_line_sample(n = 20) %>%  
  st_cast("POINT") %>% 
  st_transform(latlong) %>% 
  st_as_sf()

sites_2 <- filter(all_sites, site_ID %in% c("20170104", "20170106"))
sites_7 <- filter(all_sites, !site_ID %in% c("20170104", "20170106"))

atcs1 <- rdz_noresi[st_nearest_feature(sites_7, rdz_noresi),] %>% 
  mutate(site_ID = sites_7$site_ID)
atcs2 <- rdz_resi[st_nearest_feature(sites_2, rdz_resi),] %>% 
  mutate(site_ID = sites_2$site_ID)

atc_snap <- rbind(atcs1, atcs2)

overline_routes <- list()
overline_trimmed <- list()
atc_vars <- list()
rt <- ur[2]
for (rt in ur){

  dr <- filter(routes_all, ID == rt)
  
  r <- overline(sl = dr, attrib = "car_driver", ncores = 4) %>% 
    mutate(ID = rt)
  
  overline_routes[[rt]] <- r
  
  r_trim <- st_intersection(r, bb_buff)
  
  overline_trimmed[[rt]] <- r_trim
  
  atcs <- r_trim[st_nearest_feature(atc_snap, r_trim),] %>% 
    mutate(site_ID = str_sub(atc_snap$site_ID, 7,-1)) %>% 
    st_set_geometry(NULL)

  atc_vars[[rt]] <- atcs
  
  }

ol_routes <- do.call(rbind, overline_routes)
ol_trimmed <- do.call(rbind, overline_trimmed)
all_atcs <- do.call(rbind, atc_vars)

gg_pal <- c("#f7766c", "#d29200", "#92aa00", "#00ba37", "#00c19e", 
  "#03b8e3", "#619cff", "#da72fb", "#fe61c2")

tm3 <- tm_shape(ol_trimmed) +
  tm_lines(palette = me_pal, breaks = bks_hr,
           lwd = "car_driver",
           scale = 20,
           title.lwd = "Number of vehicles",
           alpha = 0.6,
           col = "car_driver",
           title = "Flow (number of vehicles)"
  ) +
  tm_facets(along = "ID", free.coords = FALSE)+
  tm_layout(frame = FALSE, legend.outside = TRUE)+  
  tm_scale_bar(position = c("left", "bottom"))+
  tm_shape(atc_snap)+
  tm_dots(size = 1, col = "site_ID", palette = gg_pal)

a1 <- tmap_animation(tm3, delay = 100)

library(ggplot2)
library(gganimate)

# Make a ggplot, but add frame=year: one image per year
g1 <- ggplot(all_atcs, aes(x=site_ID, y=car_driver, fill=site_ID)) + 
  geom_bar(stat='identity') +
  theme_bw()+
  # gganimate specific bits:
  transition_states(
    ID,
    transition_length = 1,
    state_length = 1
  ) +
  ease_aes('sine-in-out')

# Save at gif:
animate(g1, fps = 4)


```

Jittering routes very quickly and comparing the outputs with measured data 

Additionally an ANPR survey was also undertaken. This data can either be used as a static count point 

```{r jittered, echo = FALSE, warning=FALSE}

# baf_cent <- st_centroid(baf[1,])
# 
# routes_all$ID <- sprintf("%02d", rep(1:NROW(jitters), each=NROW(od_jittered)))
# jitters_all$ID <- sprintf("%02d", rep(1:NROW(jitters), each=NROW(od_jittered)))
# ur <- unique(routes_all$ID)



```
